DATA 621 - Business Analytics and Data Mining

### 2020-10-27 HW3/HW3_Final_AA.Rmd
Progress? Maybe?

```{r loadData, include=FALSE}
library(tidyverse)
library(tidymodels)
library(stacks)
library(kernlab)
library(vip)
library(summarytools)
library(corrplot)
library(knitr)
library(rsample) # model 1 libraries
library(caret)
library(e1071)

@@

df = read.csv(paste0(urlRemote, pathGithub, fileTrain))
df<- as_tibble(df)

eval = read.csv(paste0(urlRemote, pathGithub, fileTest))
```

# DATA EXPLORATION
## Data Description

The training data set contains 466 records summarizing attributes of various neighborhoods in the city of Boston. The response variable is coded such that it is `1` when the neighborhood’s crime rate is above the median and `0` when it is not. In all, there are 12 predictors. These include:

@@

## Data Distribution

For each predictors, we computed and drew kernel density estimate to understand their distribution. The following plots show how predictors are distributed between areas where crime rate is higher than the median, i.e. blue, and areas where crime rate is below the median, i.e. red. It is of interest to understand variables that highlight large variations between the two groups.

As a result, the density plots reveal that most is the data is not normal. some variables are heavily right skewed, while others possess multi-model distributions. More specifically, the variable predictor for average number of rooms per dwelling, i.e. `rm`, seems to be the best variable to divide the data into the two groups, since it closely mirrors a normal distribution between groups. Another interesting predictor is `zn`, i.e. proportion of residential land zoned for large lots. It has a significant positive skew (skew = 2.18). Nearly 73% of the observations (339 of 466 total) that have a value of 0. When analyzing the difference between the crime rate groups, it is possible that areas with high crime rate do not have land zoned for large lots. This suggests that suburban areas are likely to have low crime rates because there typically have large lots, whereas urban areas have a higher crime rate since lot size are smaller than 25000 square feet.

```{r density}
# density plots
variable = names(df)
df.new = df
denplot = function(i){
  df.new$x = df.new[,variable[i]]
  ggplot(df.new, aes(x, fill = factor(target))) + 
    ggtitle(sprintf("%s", variable[i])) + 
    geom_density(alpha = 0.4, show.legend = FALSE) + 
    theme(legend.position = "none",
          plot.title = element_text(size = 10, hjust = 0.5),
          axis.title.y = element_blank(),
          axis.title.x = element_blank())
}
gridExtra::grid.arrange(denplot(1),denplot(2),denplot(3),denplot(4),
                        denplot(5),denplot(6),denplot(7),denplot(8),
                        denplot(9),denplot(10),denplot(11),denplot(12),
                        layout_matrix = rbind(c(0,1,2), c(3,4,5), c(6,7,8), c(9,10,11)))
```

Further visually inspections was conducted using boxplots to highlight how the data is spread for each variable. The graphs show that some variables have a large amount of variance between each other, for example, `zn`, `rad` and `tax`. There are also quite a few values that can be considered outliers within some variables.

```{r boxplot}
# boxplots 
df.new = reshape::melt(df, id.vars= "target") %>%
  dplyr::filter(variable != "chas") %>%
  mutate(target = as.factor(target))
ggplot(df.new, aes(x = variable, y = log(value))) + 
  geom_boxplot(color = "blue", fill = "blue", alpha = 0.2, 
               notch = TRUE, notchwidth = 0.8, 
               outlier.colour = "red", outlier.fill = "red", outlier.size = 3) + 
  stat_summary(fun.y = mean, color = "red", geom = "point", shape = 16, size = 2) + 
  coord_flip() + 
  labs(title = "Boxplot of Predictor Variables")

ggplot(df.new, aes(x = variable, y = log(value))) + 
  geom_boxplot(alpha = 0.5, outlier.colour = "red", 
               outlier.fill = "red", outlier.size = 2,
               aes(fill = target)) + 
  facet_wrap( ~ variable, scales = "free") + 
  labs(title = "Boxplot of Predictor Variables by Crime Rate") 
```

## Data Correlation

By looking at a correlation matrix, boxes with blank color indicate that the correlation was not statistically significant. It is evident that many variables are moderately to highly correlated, $\rho > 0.50$. The correlogram below shows that `nox` has a largest positive correlation with `target`, whereas `dis` has a largest negative correlation with `target`.

There is a  noticeable influence that socio-economic status has on the crime rate of an area. It is well-known that crime is concentrated in disadvantaged, urban neighborhoods in the United States. The economic segregation suggests that affluent neighborhoods may be further away in terms of distance, and as a result, disadvantaged areas are more attractive to crime because the probability of success is higher even if the targets are not as profitable. Consider areas with a higher socio-economics, these areas are correlated have a lower crime rate. Moreover, industrial areas are correlated with higher levels of nitrogen oxide concentrations. This may be an indication of why these areas are less density with residents of a higher status. This trend is also seen with the `ptratio` since a higher ratio means less funding for public institutions, which is common in areas of lower status.

Lastly, it seems that the variable `chas` which indicate whether the suburb borders the Charles River have statistically insignificant correlation with almost all of the other variables except for `medv`, which is poorly positively correlated.

```{r correlation, fig.height=6, fig.width=6}
# correlations
cor.df = function(df){
  df.m = as.matrix(df)
  n = ncol(df.m)
  p.mat = matrix(NA, n, n)
  diag(p.mat) = 0
  for (i in 1:(n - 1)){
    for (j in (i + 1):n){
      tmp = cor.test(df.m[, i], df.m[, j])
      p.mat[i, j] = p.mat[j, i] = tmp$p.value
    }
  }
  colnames(p.mat) = rownames(p.mat) = colnames(df.m)
  p.mat
}
p.mat = cor.df(df)
M = cor(df)
col = colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(round(M,1), method = "color", col = col(200),  
         type = "lower", order = "hclust", 
         addCoef.col = "black",
         tl.col = "black", tl.srt = 45,
         p.mat = p.mat, sig.level = 0.01, insig = "blank", 
         diag = FALSE)
```

# DATA PREPARATION

# BUILD MODELS

First, we will split the training data into two sets. This will allow us to perform a cross validation scheme on the models to tune parameters for optimum performance.

```{r tts}

# BUILD MODELS
# train-test split
crime_df<-
  df %>% 
  mutate(target = as.factor(target)) %>% 
  mutate_if(is.integer, as.numeric)

@@

```{r baseModel}

baseModel <- glm(target~., family = binomial, data = train_df)

knitr::kable(summary(baseModel)$coefficients, digits = 3L,
             caption = 'Base Model Logistic Regression Output')

```
We can immediately see that a few variables *exceed* the 0.05 p-value threshold
for significance.

@@

### Coefficient Discussion
It is important to note that the coefficients from the model are predicting whether or not the target variable is a **1** (the crime rate is *above* the median value). Additionally, it must be noted that the numeric coefficients are relative to the range of values that the variable encompasses. What this means is that it's possible to have a coefficient that seems small when we look at the absolute magnitude, but that actually has a very strong effect when applied to the data. 
```{r model1_coeff}

# 10-fold cross validation
train_control <- trainControl(method = "cv", number = 10)

# train the model on training set
model1 <- train(target ~ zn+nox+age+dis+rad+tax+ptratio+lstat+medv,
               data = train_df,
               trControl = train_control,
               method = "glm",
               family=binomial())

knitr::kable(summary(model1)$coefficients, digits = 3L,
             caption = 'Backwards Model Regression Output')

```
Let's take a look at the *signs* of the coefficients. We can see that the coefficients for the variables `zn` and `tax` are negative. This is indicative of an inverse relationship; ie, the higher these values, the less likely the crime rate is above the median. The relationship for the zone variable aligns with the findings from our initial data exploration. However, the relationship for the tax variable does not; in fact, we saw that there was a positive correlation between the two (as the tax rate increases, so does the probability that the crime rate will be above the median).

The rest of the variables have positive coefficients. Many of these are expected. For example, in our exploratory data analysis, we noticed that the `nox` variable (nitrogen oxides concentration (parts per 10 million)) has the greatest positive correlation with the target variable. This is surprising; we hadn't thought that the nitrogen oxide concentration would have as large of an impact on the crime rate, but some quick googling shows us that there may actually be a relationship. ( [https://www.sciencedaily.com/releases/2019/10/191003114007.htm](Source) )

@@

```{r}
cv_folds<- vfold_cv(train_df,
                    v = 10, repeats = 1)

crime_recipe<- 
  recipe(target ~., data=train_df)

@@

crtl_grid<- control_stack_grid()
```

@@

```{r}
logit_specs<- 
  logistic_reg(
    penalty = tune(),
    mixture = tune()
    ) %>%
  set_engine("glm") %>%
  set_mode("classification")

svm_specs<- 
  svm_rbf(
    cost = tune(),
    rbf_sigma = tune(),
    margin = tune()
  ) %>%
  set_mode("classification") %>%
  set_engine("kernlab")
```

@@

```{r}
svm_wf<-
  crime_wf %>%
  add_model(svm_specs)

@@

logit_wf<- 
  crime_wf %>%
  add_model(logit_specs)

logit_results<-
  tune_grid(
    logit_wf,
    resamples = cv_folds,
    control = crtl_grid,
    grid = 10,
    save_pred = TRUE,
    save_workflow = FALSE
)

@@

```{r}
crime_model_stack <- 
  stacks() %>%
  add_candidates(logit_results) %>%
  add_candidates(svm_results) %>%
  blend_predictions() %>%
  fit_members()

@@

```{r}
crime_pred <- 
  test_df %>%
  bind_cols(predict(crime_model_stack, ., type = "prob"), 
            predict(crime_model_stack, ., type = "class"))
```

@@

```{r}
roc<- yardstick::roc_auc(
  crime_pred,
  truth = target,
  contains(".pred_1")
  )

@@

crime_pred %>%
  conf_mat(truth = target, estimate = .pred_class)
```

```{r}
metrics_df
```

>>>>

```{r loadData, include=FALSE}
library(tidyverse)
library(tidymodels)
library(data.table)
library(stacks)
library(kernlab)
library(vip)
library(summarytools)
library(corrplot)
library(knitr)
library(rsample) # model 1 libraries
library(caret)
library(e1071)
library(geoR)

@@

df <- read.csv(paste0(urlRemote, pathGithub, fileTrain))
nobs <- dim(df)[[1]]
DT <- as.data.table(df)
DT[, target := factor(target)]
# df <- as_tibble(df)
eval <- read.csv(paste0(urlRemote, pathGithub, fileTest))
```

# DATA EXPLORATION
## Data Description
The training data set contains 466 records summarizing attributes of various
neighborhoods in the city of Boston. The response variable is coded such that it
is `1` when the neighborhood’s crime rate is above the median and `0` when it is
not. In all, there are 12 predictors. These include:

@@

## Data Distribution
For each predictors, we computed kernel density estimators to understand their
distribution. The following plots show how predictors are distributed between
areas where the crime rate is higher than the median (blue) and areas where the
crime rate is below the median (red). It is of interest to understand variables
that highlight large variations between the two groups.

```{r density, fig.width=6}
# density plots
DT[, IDX := .I]
DTM <- melt(DT, id.vars = c('IDX', 'target'), measure.vars = 1:12,
            variable.name = 'metric', value.name = 'value')
ggplot(DTM, aes(x = value, fill = target, color = target)) +
  geom_density(alpha = 0.4, show.legend = FALSE) +
  facet_wrap(~ metric, scales = 'free') 
```

The density plots reveal that most of the data does not have a normal
distribution. There are some are heavily right tailed variables and others which
are multi-modal. The most Gaussian of the variables appears to be the one
related to the average number of rooms per dwelling: `rm`. Another
interesting predictor is `zn`, the proportion of residential land zoned for
large lots. Nearly 73% of the observations have a value of 0 with the remaining
spread widly between 1% and 100%. This may the source of possible distortion. A
third highly-skewed variable is `chas`, the indicator as to whether the lot
borders the Charles River. Out of the `r nobs` observations,
`r sum(DT$chas == 0)` do **not** border the river.

To provide another view, this time highlighting outliers, the data was analyzed
using boxplos.

```{r boxplots, fig.width=6}
# boxplots

ggplot(DTM, aes(x = metric, y = value)) + 
  geom_boxplot(color = "blue", fill = "blue", alpha = 0.2, notch = TRUE,
               notchwidth = 0.8, outlier.colour = "red", outlier.fill = "red",
               outlier.size = 3) + 
  stat_summary(fun.y = mean, color = "red", geom = "point", shape = 16,
               size = 2) + coord_flip() +
  labs(title = "Boxplot of Predictor Variables") + scale_y_log10()

ggplot(DTM, aes(x = metric, y = value)) + 
  geom_boxplot(alpha = 0.5, outlier.colour = "red", outlier.fill = "red",
               outlier.size = 2, aes(fill = target)) + 
  facet_wrap( ~ metric, scales = "free") + 
  labs(title = "Boxplot of Predictor Variables by Crime Rate") +
  scale_y_log10()
```

It is clear from the second boxplot, similar to the densities, that the
distribution of the predictor variables is different for for two outcomes. This
suggests that a model will be able to extract signal from the data.

## Data Correlation
Below is a corrgram of the data. The shape of the ellipse reflects the strength
of the correlation and the color represents the direction. Thus, the diagonal
which represents the perfect correlation of each variable with itself is a blue
diagional line. The shape and the color scheme reflect the guide at the bottom
of the image.

```{r correlation, fig.height=6, fig.width=6}
# correlations
corrplot(cor(df), method = 'ellipse', type = 'lower', order = 'hclust')
```

Looking at the corrgram, we see that there are many variables which are
moderately to highly correlated, $\left|\rho\right| > 0.50$. Of note is that
`nox` has a largest positive correlation with `target` and `dis` has the largest
negative correlation with `target`. Lastly, it seems that the variable `chas`,
which indicate whether the suburb borders the Charles River, has statistically
insignificant correlation with almost all of the other variables.

# DATA PREPARATION
There are no missing values so none need to be imputed.
Different preparation techniques may be used for differen models.
## Models 1 & 2
The variable `chas` will be removed due to having statistically insignificant\
correlation with almost all the variables in the dataset. Next, outliers will be
tempered by capping from below at the 5th percentile and from above at the 95th
percentile.

```{r, model1_2_adjust}
new_df <- subset(df, select = -chas)

# Cap all observations at their 5th and 95th percentiles
low5 <- apply(new_df, 2, quantile, prob = 0.05)
up95 <- apply(new_df, 2, quantile, prob = 0.95)
for (i in seq_along(new_df)) {
  new_df[, i] <- pmin(new_df[, i], up95[i])
  new_df[, i] <- pmax(new_df[, i], low5[i])
}
dens_ageO <- ggplot(new_df, aes(age)) + geom_density()
dens_lstatO <- ggplot(new_df, aes(lstat)) + geom_density()
dens_rmO <- ggplot(new_df, aes(rm)) + geom_density()

The Boxcox transformation will be applied to the `age`, `lstat`, and `rm`
variables.

```{r m12BC}
ageBC <- boxcoxfit(new_df$age)
lstatBC <- boxcoxfit(new_df$lstat)
rmBC <- boxcoxfit(new_df$rm)
```

The transform suggests a \(\lambda\) of `r ageBC$lambda` for `age`,
`r lstatBC$lambda` for `lstat`, and `r rmBC$lambda` for `rm`

```{r, m12applyBC}
new_df$age <- new_df$age ^ ageBC$lambda
new_df$lstat <- new_df$lstat ^ lstatBC$lambda
new_df$rm <- new_df$rm ^ rmBC$lambda
```

Below are the densities of the variables prior (left) and post (right) the
Box-Cox transform. The variables `lstat` and `rm` have benefitted from the
transform, but the strong bimodality of `age` has not been affected
significantly.

```{r, m12BC_plot}
dens_age <- ggplot(new_df, aes(age)) + geom_density()
dens_lstat <- ggplot(new_df, aes(lstat)) + geom_density()
dens_rm <- ggplot(new_df, aes(rm)) + geom_density()
grid.arrange(dens_ageO, dens_lstatO, dens_rmO, dens_age, dens_lstat, dens_rm,
             layout_matrix = cbind(c(0, 1, 2), c(3, 4, 5)))
```

## Model 3
# BUILD MODELS
## Models 1 & 2
First, we will split the training data into two sets. This will allow us to
perform a cross validation scheme on the models to tune parameters for optimum
performance.

```{r m12_tts}
# BUILD MODELS
# train-test split
crime_df <- as.tibble(df) %>%
  mutate(target = as.factor(target)) %>% 
  mutate_if(is.integer, as.numeric)

@@

```{r baseModel1}
baseModel <- glm(target~., family = binomial, data = train_df)
knitr::kable(summary(baseModel)$coefficients, digits = 3L,
             caption = 'Base Model Logistic Regression Output')
```

We can immediately see that a few variables *exceed* the 0.05 p-value threshold
for significance.

@@

### Coefficient Discussion
It is important to note that the coefficients from the model are predicting
whether or not the target variable is a **1** (the crime rate is *above* the
median value). Additionally, it must be noted that the numeric coefficients are
relative to the range of values that the variable encompasses. What this means
is that it's possible to have a coefficient that seems small when we look at the
absolute magnitude, but that actually has a very strong effect when applied to
the data.

```{r model1_coeff}
# 10-fold cross validation
train_control <- trainControl(method = "cv", number = 10)

# train the model on training set
model1 <- train(target ~ zn + nox + age + dis + rad + tax + ptratio + lstat + medv,
               data = train_df,
               trControl = train_control,
               method = "glm",
               family=binomial())
knitr::kable(summary(model1)$coefficients, digits = 3L,
             caption = 'Backwards Model Regression Output')
```

Let's take a look at the *signs* of the coefficients. We can see that the
coefficients for the variables `zn` and `tax` are negative. This is indicative
of an inverse relationship; ie, the higher these values, the less likely the
crime rate is above the median. The relationship for the zone variable aligns
with the findings from our initial data exploration. However, the relationship
for the tax variable does not; in fact, we saw that there was a positive
correlation between the two (as the tax rate increases, so does the probability
that the crime rate will be above the median).

The rest of the variables have positive coefficients. Many of these are
expected. For example, in our exploratory data analysis, we noticed that the
`nox` variable (nitrogen oxides concentration (parts per 10 million)) has the
greatest positive correlation with the target variable. This is surprising; we
hadn't thought that the nitrogen oxide concentration would have as large of an
impact on the crime rate, but some quick googling shows us that there may
actually be a relationship
([https://www.sciencedaily.com/releases/2019/10/191003114007.htm](Source)).

@@

```{r m2setup}
cv_folds <- vfold_cv(train_df, v = 10, repeats = 1)

crime_recipe <- recipe(target ~., data=train_df)

@@

crtl_grid <- control_stack_grid()
```

@@

```{r m2svm}
logit_specs <- 
  logistic_reg(
    penalty = tune(),
    mixture = tune()
    ) %>%
  set_engine("glm") %>%
  set_mode("classification")

svm_specs <- 
  svm_rbf(
    cost = tune(),
    rbf_sigma = tune(),
    margin = tune()
  ) %>%
  set_mode("classification") %>%
  set_engine("kernlab")
```

@@

```{r m2stack}
svm_wf <-
  crime_wf %>%
  add_model(svm_specs)

@@

logit_wf <- 
  crime_wf %>%
  add_model(logit_specs)

logit_results <-
  tune_grid(
    logit_wf,
    resamples = cv_folds,
    control = crtl_grid,
    grid = 10,
    save_pred = TRUE,
    save_workflow = FALSE
)
```

@@

```{r m2Blend}
crime_model_stack <- 
  stacks() %>%
  add_candidates(logit_results) %>%
  add_candidates(svm_results) %>%
  blend_predictions() %>%
  fit_members()

@@

```{r m2Predict}
crime_pred <- 
  test_df %>%
  bind_cols(predict(crime_model_stack, ., type = "prob"), 
            predict(crime_model_stack, ., type = "class"))
```

@@

```{r m2Stats}
roc<- yardstick::roc_auc(
  crime_pred,
  truth = target,
  contains(".pred_1")
  )

@@

crime_pred %>%
  conf_mat(truth = target, estimate = .pred_class)

metrics_df
```
